### 2024-02-27 paper.tex
Initial commit

\documentclass[12pt, a4paper]{article}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}
\usepackage{subfigure}

\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}
\title{When two-point correlation function fails to describe a structure...}
\author{Foo Bar}
\maketitle

\section{Introduction}
AAA BBB

\section{All the math}
\subsection{Definitions}
Suppose $A$ is some region in a Euclidean space $\mathbb{R}^n$. Let
$A_1, A_2, \dots, A_N$ be subsets of $A$ so that
\begin{align*}
  \cup_{i=1}^N A_i &= A \\
  A_i \cap A_j &= \emptyset, \ i \ne j
\end{align*}
It is said that the point $x$ is in the phase $i$ if $x \in A_i$.

Define a set of characteristic functions $\chi_i(x)$ as follows:
\begin{equation*}
  \chi_i(x) = \left\{
  \begin{array}{ll}
    1 & \quad \text{if $x \in A_i$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation*}

The two-point correlation function
$S_2^{(i)}: \mathbb{R}^n \rightarrow \mathbb{R}$ for homogeneous media can
defined in the terms of $\chi$:
\begin{equation}
  S_2^{(i)}(r) = \langle \chi_i(x) \chi_i(x+r) \rangle
  \label{eq:s2}
\end{equation}
where $\langle \dots \rangle$ means ensemble average. Its meaning is given in
the following definition:
\begin{definition}
  The two-point correlation function $S_2^{(i)}(r)$ is a probability that a
  line segment $r$ thrown randomly into $A$ has its both ends in the phase $i$.
  \label{def:s2}
\end{definition}

Two-phase media in which the region $A$ is subdivided in two adjacent phases
(usually called ``void'' and ``solid'') are of special interest for practical
applications. In this case function $S_2^{(1)}$ and $S_2^{(0)}$ has a linear
dependency \cite{Torquato_book}:
\begin{equation*}
  S_2^{(1)}(r) = S_2^{(0)}(r) + 1 - 2S_2^{(0)}(0)
\end{equation*}
In our paper we will stick to $S_2^{(1)}$ function for simplicity.

In practice the region $A$ is a $n$-dimensional array, vectors $r$ and $x$ can
have integer coefficients and $x \in A^{(1)}$ is equivalent to $A[x] = 1$
(element at the index $x$ is equal to $1$). The two-point correlation function
is then computed using fast Fourier transform.

A problem of reconstruction of $A$, i.e. properly finding its subset $A_1$, from
its two-point correlation function is often arises in porous materials
science. In this paper we approach the problem of reconstruction having a
so-called full correlation map of $A$, i.e. the values $S_2^{(1)}(x)$ for
$x \in (1 \dots 2D_1-1)\times(1 \dots 2D_2-1)\times\dots$ where
$D_1, D_2, \dots$ are dimensions of the array $A$. Here multiplication by two
and subtraction of one means that we consider two-point function calculated with
zero padding, described in the next subsection.

\subsection{Boundary conditions}
The two-point correlation function can be calculated with periodic boundary
conditions and zero padding depending on whether a line segment in
\cref{def:s2} is allowed to cross the boundary of array or not. If it is allowed
to do so, its coordinates are ``wrapped'' around the boundary, i.e. taken
$\mod D$, $D = (D_1, D_2, \dots)$ (\cref{fig:s2-computation}).
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.4\linewidth]{images/periodic.png}
    \label{fig:s2-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.4\linewidth]{images/zeros.png}
    \label{fig:s2-zeros}}
  \caption[]{Computation of the two-point correlation function at the point $2$
    for one-dimensional sequence of length 8.}
  \label{fig:s2-computation}
\end{figure*}

In this paper we will primarily discuss the two-point function calculated with
zero padding.

\subsection{Collisions in the two-point function for sequences of 0's and 1's}
When reconstructing a two-phase porous medium from its correlation function, it
is important to know to what extent a reconstruction will resemble the original
image. Consider the following definition:
\begin{definition}
  Arrays $A$ and $B$ of the same dimensionality have a collision if they have
  the same values of the two-point correlation function in all points of its
  definition, but $A \ne B$.
\end{definition}

Turns out that every array $A$ has a collision of two specific forms, which we
will discuss after describing of a mathematical framework which will help us in
the analysis.

\begin{definition}
  A Z-transform of a sequence $\{a_0, a_1, \dots, a_n\}$ is a polynomial
  $\sum_{i=0}^n a_i z^{-i}$.
\end{definition}
Z-transform is much like the discrete Fourier transform, particulary it has the
same convolution theorem:
\begin{theorem}
  If $g(z)$ and $h(z)$ are images of sequences $a$ and $b$ under Z-transform,
  then $g(z)h(z) = f(z)$ where $f(z)$ is an image of $a*b$ under Z-transform,
  where $a*b$ is a discrete convolution of $a$ and $b$:
  \begin{equation*}
    (a*b)_k = \sum_{i=0}^{n} a_i b_{k-i}
  \end{equation*}
  \label{th:conv}
\end{theorem}
In the case of one-dimensional two-phase porous media $a$ the definition
\ref{eq:s2} can be rewritten as follows
\begin{equation*}
  S_2(r) = \frac{\sum_{i}a_ia_{i+r}}{N(r)}
\end{equation*}
where $N(r)$ is a normalization factor which does not depend on $a$ and can be
omitted for simplicity. The numerator is a discrete autocorrelation which is
equal to a convolution of $a$ with ``reversed'' $a$, i.e. $i$-th element of $a$
becomes $-i$-th element of a reversed sequence. Let $f(z)$ be an image of $a$
under Z-transform. Then an image $s_2(z)$ of the sum above can be calculated
using \cref{th:conv}.
\begin{equation}
  s_2(z) = f(z)f(z^{-1})
  \label{eq:s2z}
\end{equation}
Making a substitution $t^{-1} = z + z^{-1}$ in inverting $s_2(t)$ we will get a
sequence $S_2(r) = (S_2(0), S_2(1), S_2(2), \dots)$ up to a normalization by
$N(r)$. The meaning of this sequence is a number of positive elements in $a$
distanced by $r$.
\begin{example}
  Let $a$ be a sequence $(1, 0, 1, 0, 0, 1)$. Its Z-transform is
  $f(z) = 1 + z^{-2} + z^{-5}$.
  \begin{align*}
    s_2(z) &= f(z)f(z^{-1}) \\
    &= (1 + z^{-2} + z^{-5})(1 + z^2 + z^5) \\
    &= (1 + z^2 + z^5) + (z^{-2} + 1 + z^3) + (z^{-5} + z^{-3} + 1) \\
    &= 3 + (z^2 + z^{-2}) + (z^3 + z^{-3}) + (z^5 + z^{-5})
  \end{align*}
  Thus, we have $s(t) = 3 + t^{-2} + t^{-3} + t^{-5}$. Inverting, we obtain a
  sequence $(3, 0, 1, 1, 0, 1)$, which means that $a$ has three elements in
  phase $1$, zero adjacent elements in phase $1$, one pair which is distanced by
  2 (elements with indices 0 and 2), one pair distanced by 3 (indices 2 and 5)
  and one pair distanced by 5 (indices 0 and 5).
\end{example}

By looking at \cref{eq:s2z} it becomes clear that the following two
substitutions, designated $S_k$ and $R$, do not change $s_2(z)$ and hence, the
two-point correlation function:
\begin{equation}
  \begin{aligned}
    R&: f(z) \rightarrow f(z^{-1}) \\
    S_k&: f(z) \rightarrow z^{-k}f(z)
  \end{aligned}
  \label{eq:group-elts}
\end{equation}
The set
\begin{equation}
  G = \{R, S_0, S_1, S_{-1}, \dots \}
  \label{eq:group}
\end{equation}
and a binary operation $\circ$ (composition) form a group of transforms of $a$
which do not change the two-point function. Its neutral element is $S_0$ and the
associativity and invertibility are trivial to proof.

The transform $R$ means a reversion of a sequence, i.e. writing it ``the other
way around''. The transform $S_k$ means shifting a sequence to the right by $k$
elements. These are trivial transforms which are already present in the
literature. It is clear that any two sequences $a$ and $Ta$, $T \in G$ form a
collision, so if we search for collisions, it is important to ``discard''
trivial collisions like that.

To find non-trivial collisions we firstly introduce a change of variables
$x = z^{-1}$ which is justified by a symmetry of \cref{eq:s2z}. Secondly we
introduce a notion of a reverse of a polynomial and palindrome polynomials.
\begin{definition}
  A reverse of a polynomial $f(x)$ is a polynomial $x^{\deg f}f(x^{-1})$. A
  polynomial is called a palindrome if it is equal to its reverse.
\end{definition}
Examples of palindrome polynomials include $x$, $x + 1$, $x^2 - 2x + 1$, but not
$x^3+x+1$.

Now a mechanism for finding collisions for a two-phase sequence $a$ is
described. Suppose $f(x)$ is a Z-transform of $a$ with aforementioned change of
variables. Its coefficients is therefore belong to a set $\{0, 1\}$ with the
leading and the free coefficients being equal to $1$. The assumption on the free
coefficient is taken without a loss of generality. Indeed, if the free
coefficient is zero, then a monomial $x^n$ divides $f$, and the following
construction can be applied to $\hat{f} = f / x^n$ which has the same
correlation function as $f$.

A reverse of $f(x)$ has the same degree as $f$. Suppose $f$ factors into $M$
different factors in $\mathbb{Z}[x]$:
\begin{equation*}
  f(x) = \prod_{i=1}^M f_i(x)
\end{equation*}
The image of the two-point function is then
\begin{equation*}
  s(x) = f(x)f(x^{-1}) = \prod_{i=1}^M f_i(x) \prod_{i=1}^M f_i(x^{-1})
\end{equation*}
Let now select $1 \le L < M$ non-palindrome factors
$f_{r_1}, f_{r_2}, \dots, f_{r_L}$ of $f$ and replace them with their
reverses. Obviously, the resulting correlation function won't change:
\begin{equation*}
  s(x) = \underbrace{\prod_{i=1}^L x^{\deg f_{r_i}} f_{r_i}(x^{-1}) \prod_{i=1}^{M-L} f_{o_i}(x)}_{\tilde{f}(x)}
  \underbrace{\prod_{i=1}^L x^{-\deg f_{r_i}} f_{r_i}(x) \prod_{i=1}^{M-L} f_{o_i}(x^{-1})}_{\tilde{f}(x^{-1})}
\end{equation*}
where $o_1, o_2, \dots, o_{M-L}$ run through indices of original (not replaced)
factors. Now recombining factors into $\tilde{f}(x)$ and $\tilde{f}(x^{-1})$ we
obtain a polynomial $\tilde{f}$ with the same degree as $f$ and $f(0) = \tilde{f}(0)$.
Now note that with this additional demand on the free coefficient, we can
restrict the group of transforms \cref{eq:group} to a subgroup $\{S_0, R\}$
because a non-zero shift will change the degree of $f$ which is not allowed. We
have two possible outcomes regarding $\tilde{f}$:
\begin{itemize}
\item Only one factor in $f_{r_1}, \dots, f_{r_L}$ is not a palindrome. This
  gives us $\tilde{f}(x) = f(x^{-1})$ which is a trivial collision.
\item At least two factors in $f_{r_1}, \dots, f_{r_L}$ are not palindromes. In
  this case $\tilde{f} \ne f$ and cannot be obtained from $f$ using transforms
  in the subgroup $\{S_0, R\}$.
\end{itemize}

\textcolor{blue}{It is trivial to prove that $\sum_if_i^2 = \sum_i\tilde{f}_i^2$
  where $f_i$ and $\tilde{f}_i$ are coefficients of $f$ and $\tilde{f}$
  respectively. Indeed, $f$ and $\tilde{f}$ have the same correlation function,
  and evaluating $S_2(0)$ gives us the desired result. We also make a conjecture
  that $f$ and $\tilde{f}$ have the same number of non-zero coefficients and
  thus, all coefficients of $\tilde{f}$ are either $0$ or $1$.}
\textcolor{red}{It's better to find a proof for this claim}.

Now it must be clear how to find all the non-trivial collisions of $f$. To do so
one must follow \cref{alg:collisions}.
\begin{algorithm}
  \caption{An algorithm for finding non-trivial collisions of $f$ in the
    two-point correlation function.}
  \label{alg:collisions}
  \begin{algorithmic}[1]
    \Procedure{$Collisions$}{$f$}
    \State $fs \gets factor(f) \in \mathbb{Z}[x]$
    \State $ps \gets filter(isPalindrome, fs)$
    \State $nps \gets fs \setminus d$
    \If{$length(nps) \le 1$}
    \State \textbf{return} $f$ has only trivial collisions
    \EndIf
    \State $(np:nps') \gets nps$
    \State $f' \gets np \times \prod ps$
    \State $cs \gets \text{all possible combinations of $nps'$ or reverses}$
    \State \textbf{return} $map(f'\times, cs)$
    \EndProcedure
  \end{algorithmic}
\end{algorithm}
Here ``all possible combinations of $nps'$ or reverses'' needs
clarification. For each polynomial in $nps'$ either that polynomial or its
reverse is taken in a combination of factors $c$ and $cs$ is a list of all
possible $c$'s. If $f$ has $L$ non-palindrome factors as before, we get
$2^{L-1}$ polynomials (including $f$) which have the same two-point function as
$f$.

It is clear to see, that non-trivial collisions of $f$ found using
\cref{alg:collisions} are indeed all non-trivial collisions of $f$, because the
ring $\mathbb{Z}[x, x^{-1}]$ of images of the two-point function under
Z-transform is a UFD.

\begin{example}
  Let $f = x^{20} + x^{15} + x^{13} + x^{11} + x^{10} + x^7 + x^4 + x^3 + 1 \in \mathbb{Z}[x]$.
  Factors of $f$ are $f_1 = x^{13} + x^{12} + x^{11} + x^{10} + x^4 + x^3 + x^2 + x + 1$
  and $f_2 = x^7 - x^6 + x^3 - x + 1$. We see that these factors are not
  palindromes:
  $S_{-13}R f_1 = x^{13} + x^{12} + x^{11} + x^{10} + x^9 + x^3 + x^2 + x + 1 \ne f_1$ 
  and $S_{-7}R f_2 = x^7 - x^6 + x^4 - x + 1 \ne f_2$. Hence we have 2 polynomials
  which form a non-trivial collision: $f$ and
  $g = x^7 f_1(x) f_2(x^{-1}) = x^{20} + x^{17} + x^{15} + x^{11} + x^{10} + x^8 + x^7 + x^4 + 1$.
  The two-point correlation function of these polynomials is
  $s_2(t) = 9 + 2t + 2t^2 + 4t^3 + 4t^4 + 2t^5 + 2t^6 + 4t^7 + 2t^8 + 2t^9 +
  3t^{10} + 2t^{11} + t^{12} + 2t^{13} + x^{15} + x^{16} + x^{17} + x^{20}$.
  Sequences corresponding to $f$ and $g$ are
  \begin{align*}
    & 100110010011010100001 \quad \text{and} \\
    & 100010011011000101001
  \end{align*}
  Two-point correlation function (without normalization) is
  \begin{equation*}
    (9, 2, 2, 4, 4, 2, 2, 4, 2, 2, 3, 2, 1, 2, 0, 1, 1, 1, 0, 0, 1)
  \end{equation*}
\end{example}

\subsection{Multidimensional case}
All said above works also for the multidimensional case: suppose we have
$n$-dimensional array of bits $A$, i.e. values ``0'' and ``1''. Using
Z-transform we associate a polynomial $f$ with $A$:
\begin{equation*}
  f(z_1, \dots, z_n) = \sum_{i_1=0}^{D_1 - 1} \dots \sum_{i_n=0}^{D_2 - 1}
  a_{i_1\dots i_n} z_1^{-i_1} \dots z_n^{-i_n}
\end{equation*}
where $D_1 \times \dots \times D_n$ is the dimensionality of $A$. Then an image
of the two-point function under Z-transform is
\begin{equation}
  s_2(z_1, \dots, z_n) = f(z_1, \dots, z_n)f(z_1^{-1}, \dots, z_n^{-n})
  \label{eq:s2z-md}
\end{equation}

The same group \cref{eq:group} of transforms \cref{eq:group-elts} gives rise to
trivial collisions in the two-point function, where
\begin{equation}
  \begin{aligned}
    R&: f(z_1, \dots, z_n) \rightarrow f(z_1^{-1}, \dots, z_n^{-1}) \\
    S_{k_1, \dots, k_n}&: f(z) \rightarrow z_1^{-k_1}\dots z_n^{-k_n}f(z)
  \end{aligned}
  \label{eq:group-elts-md}
\end{equation}
where $R$ means a ``reversion'' of $A$ and $S_{k_1, \dots, k_n}$ translates $A$
by a vector $(k_1, \dots, k_n)$. Now, adopting the definition of a palindrome to
the multidimensional case we can apply a change in variables $x_i = z_i^{-1}$
and the \cref{alg:collisions} to finding non-trivial collisions.

\begin{example}
  Consider a two-phase two-dimensional image \cref{fig:2d-example}. Its image
  under the Z-transform is
  $f(z_1, z_2) = z_1^{-1} + z_1^{-2} + z_2^{-1} + z_1^{-1}z_2^{-1} +
  z_1^{-2}z_2^{-1} + z_1^{-1}z_2^{-2}$. An image of the two-point correlation
  function is $s_2(t_1, t_2) = 6 + 3t_1 + t_1^2 + 3t_2 + t_2^2 + \dots$. Only
  the terms contributing to the ``axial'' two-point function are shown here for
  the sake of brevity.
\end{example}
\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.4\linewidth]{images/example-s2-image.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/example-s2.png}
  \caption[]{The unnormalized two-point function (right) calculated for a
    two-phase two-dimensional image (left). ``Ones'' are yellow, ``zeros'' are
    purple.}
  \label{fig:2d-example}
\end{figure*}

\subsection{Periodic boundary conditions}
Formula \ref{eq:s2z} can be applied to calculate the two-point function for a
sequence of the length $D$ under periodic boundary conditions if the calculation
if performed modulo $z^D - 1$. The group of transforms \cref{eq:group} does not
change the two-point function. This time, $S_k$ must be understood as a circular
shift by $k$ elements to the right.
\begin{example}
  Consider a sequence $(0, 1, 1, 1, 0, 1, 1)$ of length $7$. The image of its
  correlation function under Z-transform is
  $s_2(z) = 5 + 3(z+z^{-1}) + 2(z^2 + z^{-2}) + 2(z^3 + z^{-3}) + 2(z^4 +
  z^{-4}) + (z^5 + z^{-5})$.
  Taking $s_2(z) \mod x^7-1$ we obtain
  $5 + 3z + 3z^6 + 2z^2 + 2z^5 + 2z^3 + 2z^4 + 2z^4 + 2z^3 + z^5 + z^2 \mod z^7
  - 1 = 5 + 3z + 3z^2 + 4z^3 + 4z^4 + 3z^5 + 3z^6$. The two-point correlation
  function is then $(5, 3, 3, 4, 4, 3, 3)$.
\end{example}

Unfortunately, there is no known way to test a sequence for non-trivial
collisions in periodic boundary conditions.

\section{Probability of encountering a polynomial with non-trivial collision.}

Suppose $f \in \mathbb{Z}$ is a random polynomial with independent uniformly
distributed random coefficients in $\{0, 1\}$ with the free coefficient being
equal to $1$. It is known, that the probability of $f$ being irreducible tends
to $1$ with $d = \deg f \to \infty$ \cite{konyagin1999number}.

Breuillard et~al. \cite{breuillard2019irreducibility} give the following
approximation for this probability based on extended Riemann hypothesis:
\begin{equation}
  P(\text{$f$ is irreducible}) = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
  \label{eq:prob-irr}
\end{equation}

\textcolor{red}{How to cite a theorem in the paper?}
Another useful result is that for large enough $d$ the probability
of $f$ having factorization $\Phi \tilde{f}$ where $\tilde{f}$ is an irreducible
polynomial and $\Phi$ is a product of cyclotomic polynomials is at least
\begin{equation}
  p = 1 - 2\exp(-c(\log d)^{\beta - 2})
  \label{eq:prob-cyc-1}
\end{equation}
for some $c > 0$ and $\beta > 3$ if $\zeta_K$ has no zeros $\rho$ with
$|1 - \rho| < (\log d)^\alpha/d$, $\alpha > \beta$ for all $K = \mathbb{Q}(a)$
for any root $a$ of $f$.

A stronger bound of the probability of the same factorization with higher
reliance on the Riemann hypothesis is given by:
\begin{equation}
  p = 1 - \exp(-c \sqrt{d} / \log d)
  \label{eq:prob-cyc-2}
\end{equation}
if the Riemann hypothesis holds for $\zeta_k$ for $K = \mathbb{Q}$ and all
fields $K = \mathbb{Q}(a)$ where $a$ is a root of $f$.

Now we will prove a simple lemma which connects these results with collisions in
correlation functions.
\begin{lemma}
  Let $A$ be a sufficiently long random sequence of 0's and 1's where each
  element is chosen randomly and independently with equal probability $1/2$ with
  exception of the first and the last elements which are 1's. Then, with
  probability at least
  \begin{equation*}
    p = 1 - 2\exp(-c(\log d)^\beta)
  \end{equation*}
  for some $c > 0$, $\beta > 1$ the sequence $A$ of the length $d+1$ does not
  have non-trivial collisions in the two-point correlation function.
\end{lemma}
\begin{proof}
  Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
  $x = z^{-1}$. With probability at least $p$ it factors in $\mathbb{Z}[x]$ as
  $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
  $\tilde{f}$ is irreducible. All cyclotomic polynomials with exception of
  $\Phi_1 = x - 1$ are palindromes. $\Phi_1$ does not divide $f$ because
  $f(1) = \sum\limits_{k=0}^{d-1} f_k \ge 2$. Therefore, $f$ has at most one
  non-palindrome factor and hence $A$ does not have non-trivial collisions.
\end{proof}

For the multidimensional case, our bound is more pessimistic:
\begin{lemma}
  Let $A$ be a sufficiently big random $D-dimensional$ array of 0's and 1's
  where each element is chosen randomly and independently with equal probability
  $1/2$ with exception of the first and the last elements found on the edge
  $\mathcal{L}$, which are 1's. Then, with probability at least
  \begin{equation*}
    p = 1 - \sqrt{\frac{2}{\pi d}}
  \end{equation*}
  where $d+1$ is the length of $\mathcal{L}$, $A$ does not have non-trivial
  collisions in the two-point correlation function.
\end{lemma}
\begin{proof}
  Firstly, assume $D = 2$. Let $f(x_1, x_2) \in \mathbb{Z}[x_1][x_2]$ be an
  image of $A$ under Z-transform and change of variables $x_i = z_i^{-1}$. Let
  $\phi_l$ be a mapping $\mathbb{Z}[x_1][x_2] \rightarrow \mathbb{Z}[x_1]$ which
  ``extracts'' the leading coefficient $f_l \in \mathbb{Z}[x_1]$ from a
  bivariate polynomial $f$ and $\phi_f$ be a mapping
  $\mathbb{Z}[x_1][x_2] \rightarrow \mathbb{Z}[x_1]$ which ``extracts'' the free
  coefficient $f_0 \in \mathbb{Z}[x_1]$.

  $\phi_f$ and $\phi_l$ are homomorphisms from multiplicative monoid
  $\mathbb{Z}^*[x_1][x_2]$ to $\mathbb{Z}^*[x_1]$.
  Indeed if $g = g_0 + g_1x_2 + \dots + g_nx_2^n$ and
  $h = h_0 + h_1x_2 + \dots + h_mx_2^m$ then
  \begin{align*}
    \phi_f(g h) &= \phi_f[(g_0  h_0) + (g_0 h_1 + g_1 h_0)x_2 + \dots + (g_n
      h_m) x_2^{n+m}] \\
    &= g_0 h_0 = \phi_f(g) \phi_f(h) \qquad \text{and} \\
    \phi_l(g h) &= \phi_l[(g_0  h_0) + (g_0 h_1 + g_1 h_0)x_2 + \dots + (g_n
      h_m) x_2^{n+m}] \\
    &= g_n h_m = \phi_l(g) \phi_l(h)
  \end{align*}

  Now, without loss of generality, assume that the free (resp. leading)
  coefficient $f_k$ of $f$ has a degree $d$ and a non-zero free coefficient. If
  it is irreducible in $\mathbb{Z}[x_1]$, then $f$ is irreducible in
  $\mathbb{Z}[x_1][x_2]$. Indeed, if $f$ is not irreducible, then $f = p_1 p_2$
  for some non-constant $p_1, p_2$. Applying $\phi = \phi_f$ (resp. $\phi_l$) to
  this equality, we get $\phi(f) = f_k = \phi(p_1) \phi(p_2)$ for non-constant
  $\phi(p_1), \phi(p_2)$ which contradicts with $f_k$ being irreducible. Using
  \cref{eq:prob-irr} we obtain the least bound for the probability of $f$ being
  irreducible. If $f$ is irreducible, then it cannot have two or more
  non-palindrome factors and therefore cannot have non-trivial collisions, which
  proves the lemma.

  The proof for $D > 2$ is by recursion.
\end{proof}

Despite the probability of ``meeting'' a random polynomial which has non-trivial
collisions in the two-point function is decreasing with increase of its degree,
we can easily construct ``subtrivial'' collisions of an arbitrary (sufficiently
big) degree. Let's take two polynomials $f_1, f_2$ with coefficients being in
$\{0, 1\}$. Obviously, a polynomial $f(x) = f_1(x) f_2(x^{\deg f_1 + k})$ with
integer $k > 0$ also has its coefficients in $\{0, 1\}$.

\begin{example}
  Two sequences
  \begin{align*}
    & 1101110100001101 \quad \text{and} \\
    & 1101000011011101
  \end{align*}
  are two ``subtrivial'' collisions obtained by choosing $f_1 = f_2 = 1 + x + x^3$.
\end{example}

\section{Results}

\bibliographystyle{elsarticle-num}
\bibliography{paper}
\end{document}

### 2024-02-29 paper.tex
Results

@@
 \title{When two-point correlation function fails to describe a structure...}
-\author{Foo Bar}
+\author{Biba and Boba}
 \maketitle
@@
 $x = z^{-1}$ which is justified by a symmetry of \cref{eq:s2z}. Secondly we
-introduce a notion of a reverse of a polynomial and palindrome polynomials.
+introduce a notion of a reverse \textcolor{red}{RECIPROCAL!} of a polynomial and palindrome polynomials.
 \begin{definition}
@@
   under the Z-transform is
-  $f(z_1, z_2) = z_1^{-1} + z_1^{-2} + z_2^{-1} + z_1^{-1}z_2^{-1} +
-  z_1^{-2}z_2^{-1} + z_1^{-1}z_2^{-2}$. An image of the two-point correlation
-  function is $s_2(t_1, t_2) = 6 + 3t_1 + t_1^2 + 3t_2 + t_2^2 + \dots$. Only
-  the terms contributing to the ``axial'' two-point function are shown here for
-  the sake of brevity.
+  $f(z_1, z_2) = z_1^{-1} + z_1^{-2} + z_2^{-1} + z_1^{-1}z_2^{-1} + z_1^{-2}z_2^{-1} + z_1^{-1}z_2^{-2}$.
+  An image of the two-point correlation function is
+  $s_2(t_1, t_2) = 6 + 3t_1 + t_1^2 + 3t_2 + t_2^2 + \dots$.
+  Only the terms contributing to the ``axial'' two-point function are shown here
+  for the sake of brevity.
 \end{example}
@@
   Taking $s_2(z) \mod x^7-1$ we obtain
-  $5 + 3z + 3z^6 + 2z^2 + 2z^5 + 2z^3 + 2z^4 + 2z^4 + 2z^3 + z^5 + z^2 \mod z^7
-  - 1 = 5 + 3z + 3z^2 + 4z^3 + 4z^4 + 3z^5 + 3z^6$. The two-point correlation
-  function is then $(5, 3, 3, 4, 4, 3, 3)$.
+  $5 + 3z + 3z^6 + 2z^2 + 2z^5 + 2z^3 + 2z^4 + 2z^4 + 2z^3 + z^5 + z^2 \mod z^7 - 1 = 5 + 3z + 3z^2 + 4z^3 + 4z^4 + 3z^5 + 3z^6$.
+  The two-point correlation function is then $(5, 3, 3, 4, 4, 3, 3)$.
 \end{example}
@@
 
-A stronger bound of the probability of the same factorization with higher
-reliance on the Riemann hypothesis is given by:
-\begin{equation}
-  p = 1 - \exp(-c \sqrt{d} / \log d)
-  \label{eq:prob-cyc-2}
-\end{equation}
-if the Riemann hypothesis holds for $\zeta_k$ for $K = \mathbb{Q}$ and all
-fields $K = \mathbb{Q}(a)$ where $a$ is a root of $f$.
-
 Now we will prove a simple lemma which connects these results with collisions in
 correlation functions.
-\begin{lemma}
+\begin{theorem}
+  \label{the:sequence}
   Let $A$ be a sufficiently long random sequence of 0's and 1's where each
@@
   have non-trivial collisions in the two-point correlation function.
-\end{lemma}
+\end{theorem}
 \begin{proof}
@@
 
-For the multidimensional case, our bound is more pessimistic:
+To obtain some results for the multidimensional case, we need to prove an
+auxiliary result first.
 \begin{lemma}
-  Let $A$ be a sufficiently big random $D-dimensional$ array of 0's and 1's
-  where each element is chosen randomly and independently with equal probability
-  $1/2$ with exception of the first and the last elements found on the edge
-  $\mathcal{L}$, which are 1's. Then, with probability at least
-  \begin{equation*}
-    p = 1 - \sqrt{\frac{2}{\pi d}}
-  \end{equation*}
-  where $d+1$ is the length of $\mathcal{L}$, $A$ does not have non-trivial
-  collisions in the two-point correlation function.
+  \label{lem:factors}
+  Let $\phi$ be a homomorphism from $R[x]$ to $R'[x]$ and $\phi f \ne const$ if
+  $f \ne const$. When $f$ has no more irreducible factors (counted with their
+  repetitions) than $\phi f$.
 \end{lemma}
 \begin{proof}
-  Firstly, assume $D = 2$. Let $f(x_1, x_2) \in \mathbb{Z}[x_1][x_2]$ be an
-  image of $A$ under Z-transform and change of variables $x_i = z_i^{-1}$. Let
-  $\phi_l$ be a mapping $\mathbb{Z}[x_1][x_2] \rightarrow \mathbb{Z}[x_1]$ which
-  ``extracts'' the leading coefficient $f_l \in \mathbb{Z}[x_1]$ from a
-  bivariate polynomial $f$ and $\phi_f$ be a mapping
-  $\mathbb{Z}[x_1][x_2] \rightarrow \mathbb{Z}[x_1]$ which ``extracts'' the free
-  coefficient $f_0 \in \mathbb{Z}[x_1]$.
+  Suppose $f = \prod_{k=1}^n f_k^{m_k}$ where $f_k$ is irreducible in
+  $R[x]$. Applying $\phi$ to $f$ we get $\phi f = \prod_{k=1}^n (\phi f_k)^{m_k}$.
+  Since no $\phi f_k$ is constant, their product cannot be contracted to an
+  irreducible polynomial in $R'[x]$. Each factor $\phi f_k$ can be factored to
+  one or more irreducible polynomials in $R'[x]$ which proves the lemma.
+\end{proof}
 
-  $\phi_f$ and $\phi_l$ are homomorphisms from multiplicative monoid
-  $\mathbb{Z}^*[x_1][x_2]$ to $\mathbb{Z}^*[x_1]$.
-  Indeed if $g = g_0 + g_1x_2 + \dots + g_nx_2^n$ and
-  $h = h_0 + h_1x_2 + \dots + h_mx_2^m$ then
-  \begin{align*}
-    \phi_f(g h) &= \phi_f[(g_0  h_0) + (g_0 h_1 + g_1 h_0)x_2 + \dots + (g_n
-      h_m) x_2^{n+m}] \\
-    &= g_0 h_0 = \phi_f(g) \phi_f(h) \qquad \text{and} \\
-    \phi_l(g h) &= \phi_l[(g_0  h_0) + (g_0 h_1 + g_1 h_0)x_2 + \dots + (g_n
-      h_m) x_2^{n+m}] \\
-    &= g_n h_m = \phi_l(g) \phi_l(h)
-  \end{align*}
+Now we get our main result:
+\begin{theorem}
+  Let $A$ be a 2-dimensional $D_1 \times D_2$ array of 0's and 1's where each
+  element is chosen randomly and independently with probability $1/2$ with
+  exception of three elements $A[0, 0]$, $A[D_1 - 1, 0]$ and $A[0, D_2 - 1]$,
+  which are 1's. Then $A$ does not have non-trivial collisions in the two-point
+  correlation function with probability $p \to 1$ as $D_1, D_2 \to \infty$.
+\end{theorem}
+\begin{proof}
+  Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
+  $x = z_1^{-1}, y = z_2^{-1}$. Let $\phi f = f \mod y$ be a
+  ``partial evaluation'' homomorphism from $\mathbb{Z}[x, y]$ to
+  $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under conditions of the
+  theorem $f(0) = 1$ and $d = \deg f_x = D_2 - 1$.  According to
+  \cref{eq:prob-irr} we have $P\{\text{$f_x$ is irreducible}\} \to 1$ with
+  $d \to \infty$.
 
-  Now, without loss of generality, assume that the free (resp. leading)
-  coefficient $f_k$ of $f$ has a degree $d$ and a non-zero free coefficient. If
-  it is irreducible in $\mathbb{Z}[x_1]$, then $f$ is irreducible in
-  $\mathbb{Z}[x_1][x_2]$. Indeed, if $f$ is not irreducible, then $f = p_1 p_2$
-  for some non-constant $p_1, p_2$. Applying $\phi = \phi_f$ (resp. $\phi_l$) to
-  this equality, we get $\phi(f) = f_k = \phi(p_1) \phi(p_2)$ for non-constant
-  $\phi(p_1), \phi(p_2)$ which contradicts with $f_k$ being irreducible. Using
-  \cref{eq:prob-irr} we obtain the least bound for the probability of $f$ being
-  irreducible. If $f$ is irreducible, then it cannot have two or more
-  non-palindrome factors and therefore cannot have non-trivial collisions, which
-  proves the lemma.
-
-  The proof for $D > 2$ is by recursion.
+  We have three possible cases of factorization of $f$:
+  \begin{itemize}
+  \item $f$ is irreducible.
+  \item $f$ factors as $f(x, y) = g(x)h(y)$. In this case
+    $\phi h = h(0) = const$ and \cref{lem:factors} cannot be applied. According
+    to distribution of coefficients in $f$ the probability of this case is
+    $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$.
+  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions of
+    the theorem, $\phi g \ne const$ and $\phi h \ne const$. We have
+    $f_x = (\phi g) (\phi h)$. As we have shown, the probability of this event
+    tends to zero with $D_2 \to \infty$. Applying \cref{lem:factors}, we can say
+    the same about probability of this case.
+  \end{itemize}
+  Hence, the probability of the first case tends to one, with increase in
+  $D_1, D_2$. If $f$ is irreducible, there is no two or more non-palindrome
+  factors of $f$, hence $A$ does not have non-trivial collisions.
 \end{proof}
+This theorem also can be proved for multidimensional arrays by recursively
+applying the same argument.
 
@@
 \section{Results}
+We have written a program in Common Lisp language
+(\url{https://github.com/fatimp/ac-collisions-mt}) which searches for collisions
+in the two-point correlation functions for random sequences of 0's and 1's with
+the first and the last elements being equal to 1. The number of collisions for
+each length are present in \cref{tab:numbers}. Some examples of non-trivial
+collisions are in \cref{tab:collisions}.
+\begin{table}[!pt]
+  \centering
+  \begin{tabular}{|c|c|c||c|c|c|}
+    \hline
+    $d$ & $N_{col}$ & $N_{irred}$ & $d$ & $N_{col}$ & $N_{irred}$\\
+    \hline
+    \hline
+    25 & 20040 & 15146508 & 38 & 344 & 16269320 \\
+    \hline
+    26 & 13698 & 15400557 & 39 & 278 & 16137897 \\
+    \hline
+    27 & 10787 & 15241401 & 40 & 173 & 16454421 \\
+    \hline
+    28 & 7615 & 15656751 & 41 & 128 & 16257667 \\
+    \hline
+    29 & 5808 & 15365485 & 42 & 89 & 16565040 \\
+    \hline
+    30 & 4199 & 15883361 & 43 & 62 & 16383822 \\
+    \hline
+    31 & 3285 & 15595498 & 44 & 39 & 16634958 \\
+    \hline
+    32 & 2315 & 16023603 & 45 & 39 & 16511214 \\
+    \hline
+    33 & 1782 & 15821891 & 46 & 23 & 16630491 \\
+    \hline
+    34 & 1161 & 15988106 & 47 & 34 & 16488227 \\ 
+    \hline
+    35 & 926 & 15811952 & 48 & 20 & 16860506 \\
+    \hline
+    36 & 589 & 16356337 & 49 & 9 & 16691221 \\
+    \hline
+    37 & 471 & 16102018 & 50 & 13 & 16829791 \\
+    \hline
+  \end{tabular}
+  \caption{The number of non-trivial collisions in the two-point correlation
+    function and the number of irreducible polynomials for degrees
+    $d = [25 \dots 50]$. The degree $d$ equals to the length of a random
+    sequence minus one. The collisions were calculated by making $2 \cdot 10^7$
+    random trials for each degree (polynomials were taken with repetitions).}
+  \label{tab:numbers}
+\end{table}
+\begin{table}[!pt]
+  \centering
+  \begin{tabular}{|c|r|}
+    \hline
+    $d$ & Sequences which are non-trivial collisions \\
+    \hline
+    25 & 10000001000100000000100101 \\
+    & 10000100010010000000000101 \\
+    \hline
+    & 10111111011111101111110011 \\
+    & 11111011111101111110111001 \\
+    \hline
+    30 & 1000010000010100000000000001001 \\
+    & 1000000010000010000000000101001 \\
+    \hline
+    & 1110111101010001100101100001111 \\
+    & 1111110101001011001000111000111 \\
+    & 1111010010000111000110111010111 \\
+    & 1110011100001001010111010011111 \\
+    \hline
+    & 1110111111010101011011110111111 \\
+    & 1111110111010110101011111110111 \\
+    \hline
+    & 1001001011011110110100110100001 \\
+    & 1000010010110110110101111001001 \\
+    \hline
+    40 & 10010110000000000000000010011011001001011 \\
+    & 11010010000000000000000011011001001101001 \\
+    \hline
+    & 10111111010101111111101010011111111100011 \\
+    & 10010101111111101010111111011100011111111 \\
+    \hline
+    & 11011010010011101100011100101100110100001 \\
+    & 10010010110111001000111101001100101100011 \\
+    \hline
+  \end{tabular}
+  \caption{Examples of non-trivial collisions in the two-point correlation
+    function.}
+  \label{tab:collisions}
+\end{table}
+
+We have experimentally computed the probability of a sequence not having
+non-trivial collisions and compared it with results of \cref{the:sequence}. The
+result of comparison (taking $c = 0.148, \beta = 3.373$) is on
+\cref{fig:prob-comparison}.
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\linewidth]{images/prob.png}
+  \caption[]{Probability of a sequence of 0's and 1's with length $d+1$ and the
+    first and last elements being equal to 1 not to have non-trivial collisions
+    in the two-point correlation function: the esimate from \cref{the:sequence}
+    and a computer experiment.}
+  \label{fig:prob-comparison}
+\end{figure}
+
+For two-dimensional arrays we were able to find non-trivial collisions for array
+sizes at most $7 \times 7$ (with $10^6$ trials). One of such collisions is on
+\cref{fig:collision-2d}.
+\begin{figure*}[tp]
+  \centering
+  \includegraphics[width=0.3\linewidth]{images/collision-2d-1.png}
+  \hfill
+  \includegraphics[width=0.3\linewidth]{images/collision-2d-2.png}
+  \hfill
+  \includegraphics[width=0.3\linewidth]{images/collision-2d-s2.png}
+  \caption[]{Two 2-dimensional $7 \times 7$ arrays (left, middle), having the
+    same two-point correlation function (right).}
+  \label{fig:collision-2d}
+\end{figure*}
+
+\newpage
 

### 2024-03-05 paper.tex
Small fixes

@@
 \begin{definition}
-  Arrays $A$ and $B$ of the same dimensionality have a collision if they have
-  the same values of the two-point correlation function in all points of its
-  definition, but $A \ne B$.
+  Arrays $A$ and $B$ of the same dimensionality have a collision in the
+  two-point correlation function if they have the same values of this function
+  in all points of its definition, but $A \ne B$.
 \end{definition}
@@
 Turns out that every array $A$ has a collision of two specific forms, which we
-will discuss after describing of a mathematical framework which will help us in
-the analysis.
+will discuss after describing a mathematical framework which will help us in the
+analysis.
 
@@
   If $g(z)$ and $h(z)$ are images of sequences $a$ and $b$ under Z-transform,
-  then $g(z)h(z) = f(z)$ where $f(z)$ is an image of $a*b$ under Z-transform,
-  where $a*b$ is a discrete convolution of $a$ and $b$:
+  then $g(z)h(z) = f(z)$ where $f(z)$ is an image of $a*b$ under Z-transform
+  and $a*b$ is a discrete convolution of $a$ and $b$:
   \begin{equation*}
@@
 \end{equation}
-Making a substitution $t^{-1} = z + z^{-1}$ in inverting $s_2(t)$ we will get a
+Making a substitution $t^{-1} = z + z^{-1}$ and inverting $s_2(t)$ we will get a
 sequence $S_2(r) = (S_2(0), S_2(1), S_2(2), \dots)$ up to a normalization by
@@
 which do not change the two-point function. Its neutral element is $S_0$ and the
-associativity and invertibility are trivial to proof.
+associativity and invertibility are trivial to prove.
 
@@
 $x = z^{-1}$ which is justified by a symmetry of \cref{eq:s2z}. Secondly we
-introduce a notion of a reverse \textcolor{red}{RECIPROCAL!} of a polynomial and palindrome polynomials.
+introduce a notion of a reciprocal polynomial and palindrome polynomials.
 \begin{definition}
-  A reverse of a polynomial $f(x)$ is a polynomial $x^{\deg f}f(x^{-1})$. A
-  polynomial is called a palindrome if it is equal to its reverse.
+  A reciprocal polynomial of $f(x)$ is a polynomial $x^{\deg f}f(x^{-1})$. A
+  polynomial is called a palindrome if it is equal to its reciprocal.
 \end{definition}
@@
 
-A reverse of $f(x)$ has the same degree as $f$. Suppose $f$ factors into $M$
-different factors in $\mathbb{Z}[x]$:
+A reciprocal polynomial of $f(x)$ has the same degree as $f$. Suppose $f$
+factors into $M$ different factors in $\mathbb{Z}[x]$:
 \begin{equation*}
@@
 $f_{r_1}, f_{r_2}, \dots, f_{r_L}$ of $f$ and replace them with their
-reverses. Obviously, the resulting correlation function won't change:
+reciprocals. Obviously, the resulting correlation function won't change:
 \begin{equation*}
@@
 \begin{itemize}
-\item Only one factor in $f_{r_1}, \dots, f_{r_L}$ is not a palindrome. This
-  gives us $\tilde{f}(x) = f(x^{-1})$ which is a trivial collision.
+\item Only at most one factor in $f_{r_1}, \dots, f_{r_L}$ is not a
+  palindrome. This gives us $\tilde{f}(x) = f(x^{-1})$ which is a trivial
+  collision.
 \item At least two factors in $f_{r_1}, \dots, f_{r_L}$ are not palindromes. In
@@
   thus, all coefficients of $\tilde{f}$ are either $0$ or $1$.}
-\textcolor{red}{It's better to find a proof for this claim}.
+\textcolor{red}{It's better to find a proof for this claim, but no bad things
+  can happen, if we cannot.}
 
@@
     \Procedure{$Collisions$}{$f$}
-    \State $fs \gets factor(f) \in \mathbb{Z}[x]$
+    \State $fs \gets factor(f \in \mathbb{Z}[x])$
     \State $ps \gets filter(isPalindrome, fs)$
-    \State $nps \gets fs \setminus d$
+    \State $nps \gets fs \setminus ps$
     \If{$length(nps) \le 1$}
@@
     \State $f' \gets np \times \prod ps$
-    \State $cs \gets \text{all possible combinations of $nps'$ or reverses}$
+    \State $cs \gets \text{all possible combinations of $nps'$ or reciprocals}$
     \State \textbf{return} $map(f'\times, cs)$
@@
 \end{algorithm}
-Here ``all possible combinations of $nps'$ or reverses'' needs
+Here ``all possible combinations of $nps'$ or reciprocals'' needs
 clarification. For each polynomial in $nps'$ either that polynomial or its
-reverse is taken in a combination of factors $c$ and $cs$ is a list of all
+reciprocal is taken in a combination of factors $c$ and $cs$ is a list of all
 possible $c$'s. If $f$ has $L$ non-palindrome factors as before, we get
@@
 
-The same group \cref{eq:group} of transforms \cref{eq:group-elts} gives rise to
-trivial collisions in the two-point function, where
+The same group \cref{eq:group} of transforms \cref{eq:group-elts}:
 \begin{equation}
@@
 \end{equation}
-where $R$ means a ``reversion'' of $A$ and $S_{k_1, \dots, k_n}$ translates $A$
-by a vector $(k_1, \dots, k_n)$. Now, adopting the definition of a palindrome to
-the multidimensional case we can apply a change in variables $x_i = z_i^{-1}$
-and the \cref{alg:collisions} to finding non-trivial collisions.
+gives rise to trivial collisions in the two-point function, where $R$ means a
+``reversion'' of $A$ and $S_{k_1, \dots, k_n}$ translates $A$ by a vector
+$(k_1, \dots, k_n)$. Now, adopting the definition of a palindrome to the
+multidimensional case we can apply a change in variables $x_i = z_i^{-1}$ and
+the \cref{alg:collisions} to finding non-trivial collisions.
 
@@
 
-Suppose $f \in \mathbb{Z}$ is a random polynomial with independent uniformly
+Suppose $f \in \mathbb{Z}[x]$ is a polynomial with independent uniformly
 distributed random coefficients in $\{0, 1\}$ with the free coefficient being
@@
   $f \ne const$. When $f$ has no more irreducible factors (counted with their
-  repetitions) than $\phi f$.
+  multiplicity) than $\phi f$.
 \end{lemma}
@@
     $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$.
-  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions of
-    the theorem, $\phi g \ne const$ and $\phi h \ne const$. We have
+  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions on
+    the coefficients $\deg_x g > 0$ and $\deg_x h > 0$, therefore
+    $\phi g \ne const$ and $\phi h \ne const$. We have
     $f_x = (\phi g) (\phi h)$. As we have shown, the probability of this event

### 2024-03-05 paper.tex
Improve the main result

@@
   element is chosen randomly and independently with probability $1/2$ with
-  exception of three elements $A[0, 0]$, $A[D_1 - 1, 0]$ and $A[0, D_2 - 1]$,
-  which are 1's. Then $A$ does not have non-trivial collisions in the two-point
-  correlation function with probability $p \to 1$ as $D_1, D_2 \to \infty$.
+  exception of two elements $A[0, 0]$ and $A[D_1 - 1, 0]$, which are 1's. Then
+  $A$ does not have non-trivial collisions in the two-point correlation function
+  with probability at least
+  \begin{equation}
+    p = 1 - \sqrt{\frac{2}{\pi (D_1 - 1)}}
+    \label{eq:main-result}
+  \end{equation}
 \end{theorem}
@@
   Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
-  $x = z_1^{-1}, y = z_2^{-1}$. Let $\phi f = f \mod y$ be a
-  ``partial evaluation'' homomorphism from $\mathbb{Z}[x, y]$ to
-  $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under conditions of the
-  theorem $f(0) = 1$ and $d = \deg f_x = D_2 - 1$.  According to
-  \cref{eq:prob-irr} we have $P\{\text{$f_x$ is irreducible}\} \to 1$ with
-  $d \to \infty$.
+  $x = z_1^{-1}, y = z_2^{-1}$ with $\deg_x f = d$, where $d = D_1 - 1$. Let
+  $\phi f = f \mod y$ be a ``partial evaluation'' homomorphism from
+  $\mathbb{Z}[x, y]$ to $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under
+  conditions of the theorem $f(0) = 1$ and $\deg f_x = d$.  According to
+  \cref{eq:prob-irr} we have
+  \begin{equation*}
+    p\{\text{$f_x$ is irreducible}\} = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
+  \end{equation*}
+  Now we need to prove a similar result for $f$.
 
@@
     to distribution of coefficients in $f$ the probability of this case is
-    $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$.
+    $p_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$. For any fixed integer $D_2 > 1$
+    we have $p_{split} = o(D_1^{-1})$. For $D_2 = 1$ we have $f = f_x$ and
+    \cref{eq:prob-irr} is applicable directly to $f$.
   \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions on
@@
     $\phi g \ne const$ and $\phi h \ne const$. We have
-    $f_x = (\phi g) (\phi h)$. As we have shown, the probability of this event
-    tends to zero with $D_2 \to \infty$. Applying \cref{lem:factors}, we can say
-    the same about probability of this case.
+    $f_x = (\phi g) (\phi h)$. Applying \cref{lem:factors}, we can say that
+    $f$ is irreducible with probability at least \cref{eq:main-result}.
   \end{itemize}
-  Hence, the probability of the first case tends to one, with increase in
-  $D_1, D_2$. If $f$ is irreducible, there is no two or more non-palindrome
-  factors of $f$, hence $A$ does not have non-trivial collisions.
+  Hence $f$ is irreducible with probability at least \cref{eq:main-result}. If
+  $f$ is irreducible, there is no two or more non-palindrome factors of $f$,
+  hence $A$ does not have non-trivial collisions.
 \end{proof}
-This theorem also can be proved for multidimensional arrays by recursively
-applying the same argument.
+Surely, $D_1$ can be replaced with $D_2$ in \cref{eq:main-result} if
+$A[0, D_2 - 1] = 1$. This theorem also can be proved for multidimensional arrays
+by recursively applying the same argument.
 

### 2024-03-06 paper.tex
Be more explicit about the Riemann hypothesis

@@
 Breuillard et~al. \cite{breuillard2019irreducibility} give the following
-approximation for this probability based on extended Riemann hypothesis:
+approximation for this probability if the Riemann hypothesis holds for the
+Dedekind zeta function $\zeta_K$ for all number fields of the form
+$K = \mathbb{Q}(a)$, where $a$ is a root of $f$.
 \begin{equation}
@@
   element is chosen randomly and independently with probability $1/2$ with
-  exception of two elements $A[0, 0]$ and $A[D_1 - 1, 0]$, which are 1's. Then
-  $A$ does not have non-trivial collisions in the two-point correlation function
-  with probability at least
+  exception of two elements $A[0, 0]$ and $A[D_1 - 1, 0]$, which are
+  1's. Suppose the Riemann hypothesis for the Dedekind zeta function $\zeta_K$
+  holds for all fields $K = \mathbb{Q}(a)$ where $a$ is any root of $f_x(x)$ and
+  $f_x(x)$ is an image of the slice $A[0:D_1 -1, 0]$ under Z-transform and a
+  change in variables $x = z^{-1}$. Then $A$ does not have non-trivial
+  collisions in the two-point correlation function with probability at least
   \begin{equation}
@@
   $\phi f = f \mod y$ be a ``partial evaluation'' homomorphism from
-  $\mathbb{Z}[x, y]$ to $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under
-  conditions of the theorem $f(0) = 1$ and $\deg f_x = d$.  According to
-  \cref{eq:prob-irr} we have
+  $\mathbb{Z}[x, y]$ to $\mathbb{Z}[x]$. Oviously,
+  $f_x(x) = \phi f = f(x,0)$. Under conditions of the theorem $f(0) = 1$ and
+  $\deg f_x = d$.  According to \cref{eq:prob-irr} we have
   \begin{equation*}

### 2024-03-06 paper.tex
Capitalize probabilities

@@
 \begin{equation}
-  p = 1 - 2\exp(-c(\log d)^{\beta - 2})
+  P = 1 - 2\exp(-c(\log d)^{\beta - 2})
   \label{eq:prob-cyc-1}
@@
   \begin{equation*}
-    p = 1 - 2\exp(-c(\log d)^\beta)
+    P = 1 - 2\exp(-c(\log d)^\beta)
   \end{equation*}
@@
   Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
-  $x = z^{-1}$. With probability at least $p$ it factors in $\mathbb{Z}[x]$ as
+  $x = z^{-1}$. With probability at least $P$ it factors in $\mathbb{Z}[x]$ as
   $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
@@
   \begin{equation}
-    p = 1 - \sqrt{\frac{2}{\pi (D_1 - 1)}}
+    P = 1 - \sqrt{\frac{2}{\pi (D_1 - 1)}}
     \label{eq:main-result}
@@
   \begin{equation*}
-    p\{\text{$f_x$ is irreducible}\} = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
+    P\{\text{$f_x$ is irreducible}\} = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
   \end{equation*}
@@
     to distribution of coefficients in $f$ the probability of this case is
-    $p_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$. For any fixed integer $D_2 > 1$
-    we have $p_{split} = o(D_1^{-1})$. For $D_2 = 1$ we have $f = f_x$ and
+    $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$. For any fixed integer $D_2 > 1$
+    we have $P_{split} = o(D_1^{-1})$. For $D_2 = 1$ we have $f = f_x$ and
     \cref{eq:prob-irr} is applicable directly to $f$.

### 2024-03-07 paper.tex
Remove wrong assumption on periodicity
Inverting -> inverting Z-transform

@@
 (usually called ``void'' and ``solid'') are of special interest for practical
-applications. In this case function $S_2^{(1)}$ and $S_2^{(0)}$ has a linear
-dependency \cite{Torquato_book}:
-\begin{equation*}
-  S_2^{(1)}(r) = S_2^{(0)}(r) + 1 - 2S_2^{(0)}(0)
-\end{equation*}
-In our paper we will stick to $S_2^{(1)}$ function for simplicity.
+applications. In our paper we will stick to $S_2^{(1)}$ function for simplicity.
 
@@
 \end{equation}
-Making a substitution $t^{-1} = z + z^{-1}$ and inverting $s_2(t)$ we will get a
-sequence $S_2(r) = (S_2(0), S_2(1), S_2(2), \dots)$ up to a normalization by
-$N(r)$. The meaning of this sequence is a number of positive elements in $a$
-distanced by $r$.
+Making a substitution $t^{-1} = z + z^{-1}$ and inverting Z-transform for
+$s_2(t)$ we will get a sequence $S_2(r) = (S_2(0), S_2(1), S_2(2), \dots)$ up to
+a normalization by $N(r)$. The meaning of this sequence is a number of positive
+elements in $a$ distanced by $r$.
 \begin{example}
@@
   \end{align*}
-  Thus, we have $s(t) = 3 + t^{-2} + t^{-3} + t^{-5}$. Inverting, we obtain a
-  sequence $(3, 0, 1, 1, 0, 1)$, which means that $a$ has three elements in
-  phase $1$, zero adjacent elements in phase $1$, one pair which is distanced by
-  2 (elements with indices 0 and 2), one pair distanced by 3 (indices 2 and 5)
-  and one pair distanced by 5 (indices 0 and 5).
+  Thus, we have $s(t) = 3 + t^{-2} + t^{-3} + t^{-5}$. Inverting Z-transform, we
+  obtain a sequence $(3, 0, 1, 1, 0, 1)$, which means that $a$ has three
+  elements in phase $1$, zero adjacent elements in phase $1$, one pair which is
+  distanced by 2 (elements with indices 0 and 2), one pair distanced by 3
+  (indices 2 and 5) and one pair distanced by 5 (indices 0 and 5).
 \end{example}

### 2024-03-07 paper.tex
Describe why we do a substitution t^{-1} = z + z^{-1}

@@
 \end{equation}
-Making a substitution $t^{-1} = z + z^{-1}$ and inverting Z-transform for
-$s_2(t)$ we will get a sequence $S_2(r) = (S_2(0), S_2(1), S_2(2), \dots)$ up to
-a normalization by $N(r)$. The meaning of this sequence is a number of positive
-elements in $a$ distanced by $r$.
+Because $S_2(r) = S_2(-r)$ it is convenient to make a substitution $t^{-1} = z +
+z^{-1}$ and consider a function of a non-negative argument. Inverting
+Z-transform for $s_2(t)$ we will get a sequence
+$S_2(r) = (S_2(0), S_2(1), S_2(2), \dots)$ up to a normalization by $N(r)$. The
+meaning of this sequence is a number of positive elements in $a$ distanced by
+$r$.
 \begin{example}

### 2024-03-07 paper.tex
Fix some statements about non-palindrome factors

@@
 \end{equation*}
-Let now select $1 \le L < M$ non-palindrome factors
-$f_{r_1}, f_{r_2}, \dots, f_{r_L}$ of $f$ and replace them with their
-reciprocals. Obviously, the resulting correlation function won't change:
+Let now select all non-palindrome factors $f_{r_1}, f_{r_2}, \dots, f_{r_L}$ of
+$f$, $0 \le L \le M$ and replace them with their reciprocals. Obviously, the
+resulting correlation function won't change:
 \begin{equation*}
@@
 \begin{itemize}
-\item Only at most one factor in $f_{r_1}, \dots, f_{r_L}$ is not a
-  palindrome. This gives us $\tilde{f}(x) = f(x^{-1})$ which is a trivial
+\item Only at most one factor of $\tilde{f}$ is not a palindrome
+  ($L \le 1$). This gives us $\tilde{f}(x) = f(x^{-1})$ which is a trivial
   collision.
-\item At least two factors in $f_{r_1}, \dots, f_{r_L}$ are not palindromes. In
+\item At least two factors of $\tilde{f}$ are not palindromes ($L > 1$). In
   this case $\tilde{f} \ne f$ and cannot be obtained from $f$ using transforms

### 2024-03-07 paper.tex
S₀ → Id

@@
 \begin{equation}
-  G = \{R, S_0, S_1, S_{-1}, \dots \}
+  G = \{R, Id, S_1, S_{-1}, \dots \}
   \label{eq:group}
@@
 and a binary operation $\circ$ (composition) form a group of transforms of $a$
-which do not change the two-point function. Its neutral element is $S_0$ and the
-associativity and invertibility are trivial to prove.
+which do not change the two-point function. Its neutral element $Id = S_0$ and
+the associativity and invertibility are trivial to prove.
 
@@
 Now note that with this additional demand on the free coefficient, we can
-restrict the group of transforms \cref{eq:group} to a subgroup $\{S_0, R\}$
+restrict the group of transforms \cref{eq:group} to a subgroup $\{Id, R\}$
 because a non-zero shift will change the degree of $f$ which is not allowed. We
@@
   this case $\tilde{f} \ne f$ and cannot be obtained from $f$ using transforms
-  in the subgroup $\{S_0, R\}$.
+  in the subgroup $\{Id, R\}$.
 \end{itemize}

### 2024-03-07 paper.tex
Cite theorems properly

@@
 
-Breuillard et~al. \cite{breuillard2019irreducibility} give the following
-approximation for this probability if the Riemann hypothesis holds for the
-Dedekind zeta function $\zeta_K$ for all number fields of the form
+Breuillard et~al. \cite[Corollary~4]{breuillard2019irreducibility} give the
+following approximation for this probability if the Riemann hypothesis holds for
+the Dedekind zeta function $\zeta_K$ for all number fields of the form
 $K = \mathbb{Q}(a)$, where $a$ is a root of $f$.
@@
 
-\textcolor{red}{How to cite a theorem in the paper?}
 Another useful result is that for large enough $d$ the probability
@@
 $|1 - \rho| < (\log d)^\alpha/d$, $\alpha > \beta$ for all $K = \mathbb{Q}(a)$
-for any root $a$ of $f$.
+for any root $a$ of $f$ \cite[Theorem~5]{breuillard2019irreducibility}.
 

### 2024-03-12 paper.tex
Prove blue text

@@
 
-\textcolor{blue}{It is trivial to prove that $\sum_if_i^2 = \sum_i\tilde{f}_i^2$
-  where $f_i$ and $\tilde{f}_i$ are coefficients of $f$ and $\tilde{f}$
-  respectively. Indeed, $f$ and $\tilde{f}$ have the same correlation function,
-  and evaluating $S_2(0)$ gives us the desired result. We also make a conjecture
-  that $f$ and $\tilde{f}$ have the same number of non-zero coefficients and
-  thus, all coefficients of $\tilde{f}$ are either $0$ or $1$.}
-\textcolor{red}{It's better to find a proof for this claim, but no bad things
-  can happen, if we cannot.}
+\textcolor{red}{The idea of proof for this lemma is a contribution of Kolya and
+  Oleg}.
+Now the following lemma shows that $\tilde{f}$ is a polynomial with coefficients
+being equal to 0 or 1:
+\begin{lemma}
+  Let $f \in \mathbb{Z}[x]$ is a polynomial with coefficients in $\{0, 1\}$
+  which factors as $f(x) = \prod_{i=1}^n f_i(x)$. A polynomial
+  $\tilde{f}(x) = \prod_k f_k(x) \prod_l x^{\deg f_l} f_l(x^{-1})$ where $k$
+  runs through some subset $K$ of $N = \{1, \dots, n\}$ and $l$ runs through
+  $N \setminus K$ has also its coefficients in $\{0, 1\}$.
+\end{lemma}
+\begin{proof}
+  In this proof we assume $f(0) = 1$ for simplicity.
+
+  Suppose $n = 1$. Then $\tilde{f}$ is either the reciprocal of $f$ or $f$ itself.
+
+  Suppose $n = 2$ and write $f = gh$. In two trivial cases when $K = \emptyset$
+  or $K = N$ we have $\tilde{f}$ to be equal either the reciprocal of $f$ or
+  $f$. Let $\tilde{f}(x) = \tilde{g}(x)h(x)$,
+  $\tilde{g}(x) = x^{\deg g}g(x^{-1})$. As shown above, $f$ and $\tilde{f}$ have
+  the same two-point correlation functions, i.e.
+  $s_f(t) = s_{\tilde{f}}(t)$. Comparing free coefficients of $s_f(t)$ and
+  $s_{\tilde{f}}(t)$ we obtain the following relation for coefficients of $f$:
+  \begin{equation}
+    \sum_{k=0}^d f_k^2 = \sum_{k=0}^d \tilde{f}_k^2
+    \label{eq:squares-equal}
+  \end{equation}
+  where $d = \deg f = \deg \tilde{f}$. Now we have this chain of equalities
+  \begin{equation}
+    \begin{aligned}
+    \sum_{k=0}^d f_k = f(1) &= g(1) h(1) = \sum_{k=0}^{d_g} g_k \sum_{k=0}^{d_h} h_k \\
+    = \sum_{k=0}^{d_g} g_{d_g-k} \sum_{k=0}^{d_h} h_k &= \tilde{g}(1) h(1) =
+    \tilde{f}(1) = \sum_{k=0}^d \tilde{f}_k
+    \end{aligned}
+    \label{eq:coeff-equal}
+  \end{equation}
+
+  For any $m \in \mathbb{Z} \setminus \{0, 1\}$ we have $m^2 > m$. Taking into
+  account \cref{eq:coeff-equal} and letting any coefficient of $\tilde{f}$ be
+  in set $\mathbb{Z} \setminus \{0, 1\}$ we get a contradiction with
+  \cref{eq:squares-equal}. This contradiction proves this case.
+
+  Now suppose $n > 2$. We can construct products
+  $\tilde{g}(x) = \prod_l x^{\deg f_l }f_l(x^{-1})$ and $h(x) = \prod_k f_k(x)$
+  and continue with the proof for $n = 2$.
+\end{proof}
 

### 2024-03-12 paper.tex
Kinda intro + abstract

@@
 
+\begin{abstract}
+  In this paper the two-point correlation function $S_2$ for arrays which
+  elements are either 0 or 1 is studied and mechanisms of formation of
+  collisions in $S_2$ are determined. A collision in $S_2$ is two different
+  arrays which result in the same two-point correlation function. It is shown
+  that there are only two trivial collisions obtained by simple transformations
+  of the input array, when dimensions of the input array coincide with
+  dimensions of its bounding box. It is also shown that some input arrays have
+  non-trivial collisions which cannot be obtained by application of simple
+  transformations. The algorithm for finding non-trivial collisions for any
+  particular input array is shown in the paper and it is stated that the
+  probability of finding non-trivial collision by randomly picking the input
+  array tends to zero with increase in the input array's dimensions.
+\end{abstract}
+
 \section{Introduction}
-AAA BBB
+In this paper we designate polynomials either $f$ or $f(x)$. The latter
+designation is usually used when it's important to know the variable or to
+differentiate a polynomial $f_k(x)$ from a coefficient $f_k$.
 

### 2025-01-10 paper.tex
Small style fix
Some another document class
Reformulate proof
Do not ommit mention of Dedekind zeta function

@@
-\documentclass[12pt, a4paper]{article}
+\documentclass[preprint,amsmath,amssymb,aps,pre,showkeys,showpacs]{revtex4-1}
 \usepackage[english]{babel}
@@
 \usepackage[T1]{fontenc}
-\usepackage{amsmath}
-\usepackage{amssymb}
 \usepackage{amsthm}
@@
 \title{When two-point correlation function fails to describe a structure...}
-\author{Biba and Boba}
-\maketitle
+\author{Biba}
+\author{Boba}
 
@@
 \end{abstract}
+
+\maketitle
 
@@
   in set $\mathbb{Z} \setminus \{0, 1\}$ we get a contradiction with
-  \cref{eq:squares-equal}. This contradiction proves this case.
+  \cref{eq:squares-equal}. This contradiction proves the case $n = 2$.
 
@@
 one must follow \cref{alg:collisions}.
-\begin{algorithm}
+\begin{algorithm}[H]
   \caption{An algorithm for finding non-trivial collisions of $f$ in the
@@
   correlation function under Z-transform is
-  $s_2(z) = 5 + 3(z+z^{-1}) + 2(z^2 + z^{-2}) + 2(z^3 + z^{-3}) + 2(z^4 +
-  z^{-4}) + (z^5 + z^{-5})$.
+  $s_2(z) = 5 + 3(z+z^{-1}) + 2(z^2 + z^{-2}) + 2(z^3 + z^{-3}) + 2(z^4 + z^{-4}) + (z^5 + z^{-5})$.
   Taking $s_2(z) \mod x^7-1$ we obtain
@@
   \begin{equation*}
-    P = 1 - 2\exp(-c(\log d)^\beta)
+    P = 1 - 2\exp(-c(\log d)^{\beta-2})
   \end{equation*}
-  for some $c > 0$, $\beta > 1$ the sequence $A$ of the length $d+1$ does not
-  have non-trivial collisions in the two-point correlation function.
+  for some $c > 0$, $\beta > 3$ the sequence $A$ of the length $d+1$ does not
+  have non-trivial collisions in the two-point correlation function if $\zeta_K$
+  has no zeros $\rho$ with $|1 - \rho| < (\log d)^\alpha/d$, $\alpha > \beta$
+  for all $K = \mathbb{Q}(a)$ where $a$ is a root of $f(x)$ which is an image of
+  $A$ under Z-transform and a change of variables $x = z^{-1}$.
 \end{theorem}
 \begin{proof}
-  Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
-  $x = z^{-1}$. With probability at least $P$ it factors in $\mathbb{Z}[x]$ as
+  With probability at least $P$, $f \in \mathbb{Z}[x]$ factors as
   $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
@@
     $\phi g \ne const$ and $\phi h \ne const$. We have
-    $f_x = (\phi g) (\phi h)$. Applying \cref{lem:factors}, we can say that
-    $f$ is irreducible with probability at least \cref{eq:main-result}.
+    $f_x = (\phi g) (\phi h)$. Applying \cref{lem:factors}, we can say that this
+    is the case with probability less that $1 - P$.
   \end{itemize}
@@
 
-\newpage
-
-\bibliographystyle{elsarticle-num}
 \bibliography{paper}
